Reproducible Research: Peer Assessment 1
========================================================

## Setting working directory and packages
```{r}
setwd("~/Documents/projects/data_science_specialization/reproducible_research/")
library(data.table)
```

## Loading and preprocessing the data
### Read the file
```{r}
        data<-read.csv("activity.csv")
```
### Ignore NA records under steps columns
```{r}
        step.per.day<-data[!is.na(data$steps),]
        step.per.day<-as.data.table(step.per.day)
```

## What is the mean total number of steps taken per day?
### Total number of steps per day
```{r}
        total.step.per.day<-step.per.day[,sum(steps),by=date]
```
### Histogram of total steps per day
```{r}
        hist(total.step.per.day$V1,xlab="Steps",main="Histogram of Steps per day")
```
### Mean and median of total number of steps taken per day
```{r}        
        summary(total.step.per.day$V1)[c(3,4)]        
```

## What is the average daily activity pattern?
### Average number of steps taken, averaged across all days (y) vs interval(x)
```{r}
        average.steps<-step.per.day[,list(avg.steps=ave(steps)),by=interval]
        library(lattice)
        xyplot(avg.steps~interval,data=average.steps,type="l")
```
### Which 5 minute interval, on average across all the days in the dataset, contains maximum number of steps?
```{r}
        unique(average.steps[avg.steps==max(avg.steps)])
```

## Imputing missing values
### Total number of rows with NAs
```{r}
        data<-as.data.table(data)
        dim(data[is.na(steps)])[1]
```
### Strategy to fill in missing data; mean/median for that day or mean for that 5-minute interval
```{r}
        no.mis.data<-merge(data,unique(average.steps),by="interval",all.x=T)
        no.mis.data$complete.steps<-ifelse(is.na(no.mis.data$steps),no.mis.data$avg.steps,no.mis.data$steps)
        setnames(no.mis.data,c("steps","complete.steps"),c("old.steps","steps"))        
```
       
## From the no "NA" data, plot histogram and compare if mean, median of total number of steps taken per day is different than what was observed before, when "NA" containing data was removed
### Total number of steps per day
```{r}
        total.step.per.day.no.mis.data<-no.mis.data[,sum(steps),by=date]
```        
### Histogram of total steps per day
```{r}        
        hist(total.step.per.day.no.mis.data$V1,xlab="Steps",main="Histogram of Steps per day; NAs replaced with mean/interval")
```
### Mean and median of total number of steps taken per day
```{r}
        summary(total.step.per.day.no.mis.data$V1)[c(3,4)]
```        
### If mean median (without NA and NA replaced by mean steps by interval) differ 
```{r}        
        mm.no.na.total.step.per.day<-summary(total.step.per.day$V1)[c(3,4)]
        mm.na.total.step.per.day<-summary(total.step.per.day.no.mis.data$V1)[c(3,4)]
        ifelse(mm.no.na.total.step.per.day[1]==mm.na.total.step.per.day[1],
               "No change in Median total number of steps taken per day: [NA steps and associated records removed] IS EQUAL to [NAs replaced by mean steps by interval]",
               "Change in Median total number of steps taken per day: [NA steps and associated records removed] IS NOT EQUAL to [NAs replaced by mean steps by interval]")
        ifelse(mm.no.na.total.step.per.day[2]==mm.na.total.step.per.day[2],
               "No change in Mean total number of steps taken per day: [NA steps and associated records removed] IS EQUAL to [NAs replaced by mean steps by interval]",
               "Change in Mean total number of steps taken per day: [NA steps and associated records removed] IS NOT EQUAL to [NAs replaced by mean steps by interval]")
```

## Are there differences in activity patterns between weekdays and weekends?
```{r,results='hide'}
average.steps.no.na<-no.mis.data[,"daytype":=weekdays(as.Date(date))]
average.steps.no.na[,"WKvsWEN":=ifelse(grepl("^[S]",daytype),"weekend","weekday")]
average.steps.no.na[,"avg.WKvsWEN.interval":=ave(steps),by=c("WKvsWEN","interval")]
```
```{r}
xyplot(avg.WKvsWEN.interval~interval|WKvsWEN,data=average.steps.no.na,type="l",main="Avg. number of steps: by(WKd vs WKend + interval)",ylab="Avg. number of steps",layout=c(1,2))
```
## Thank you for your time reviewing my code
